Simulations of Ballot Polling Risk-Limiting Audits

### 2022-10-12 providence/sections/adversary.tex
toy example

To illustrate the practical implication of this property, we consider a toy example: an RLA of a two-candidate contest with margin $1\%$ and risk limit $10\%$. 
Suppose we wish to achieve a conditional stopping probability $0.9$ in each round of the audit. For \Providence, we can compute a new round size for each round based on the previous samples, but for \Minerva we need to commit to a round schedule in advance. So for \Minerva we choose a first round size which gives a $0.9$ probability of stopping, and then we set the subsequent marginal round sizes to be $1.5$ times each previous round size. This multiplier of $1.5$ is known to give a probability of stopping roughly $0.9$ in the second round if the first round size has probability of stopping $0.9$. Audit software ARLO, used by many states, commits to this round schedule by default. Both audits therefore begin with first round size of $17,272$ to give a $0.9$ probability of stopping, and both will stop in the first round if the sample contains at least $8,725$ ballots for the winner. We now consider two cases for which the audit proceeds to a second round.

In one case there are $8,724$ votes for the winner in the sample, just one fewer than the minimum needed to meet the risk limit. In the \Minerva audit, we are already committed to a second round size of $43,180$ which, given the nearly-passing sample of the first round is higher than necessary, achieving a stopping probability in the second round of $.954$. The \Providence audit samples more than $9,000$ fewer ballots with a round size of $34,078$, achieving the desired $0.9$ probability of stopping.

In a less lucky sample, the winner recieves $8,637$ ballots, few more than the loser recieves. In the \Minerva audit, we again have to use a second round size of $43,180$, but now this round size only achieves a $0.727$ probability of stopping, significantly less than the desired $0.9$. Again, the \Providence audit can scale up the second round size according to the first sample and achieve the desired $0.9$ probability of stopping with $58,007$ ballots.

@@

\subsection{Efficiency}
\begin{lemma}
% For any round schedule %$[n_1, \ldots, n_r]$
For any risk-limit $\alpha \in (0, 1)$, for any margin
and for any round schedule $[n_1, \ldots, n_r]$, 
 \Providence RLA is more efficient than \R \BRAVO.
\end{lemma}
\begin{proof}
 Let $[n_1, \ldots, n_r]$ be a round schedule, and  lets assume that \R \BRAVO audit stopped exactly at
 round $r$, 
 after observing $k_1, \ldots, k_r$ ballots for the announced winner in each round respectively. This means that
 the following conditions hold:
 \begin{enumerate}
  \item for all $j = 1, \ldots, r-1$: 
%   $k_j < k^{p_a, p_0, \alpha}_{min, \BRAVO, n_j}$,
  $\frac{P[{K_j = k_{j}} | H_a, n_j]}{P[K_j = {k_{j}} | H_0, n_j]} < \frac{1}{\alpha}$,
  \item for $j = r$ it holds: % $k_r \geq k^{p_a, p_0, \alpha}_{min, \BRAVO, n_r}.$
  $\frac{P[{K_r = k_{r}} | H_a, n_r]}{P[K_r = {k_{r}} | H_0, n_r]} \geq \frac{1}{\alpha}$,
 \end{enumerate}

 From Lemma~\ref{lemma:tau1_increasing} 
 \[
 \tau_1(k, p_a, p_0, n_r) = 
  \frac{P[K_r = {k} | H_a, n_r]}{P[K_r = {k} | H_0, n_r]}
 \]
is a stricly increasing function of $k$ (for $0 < p_0, p_a < 1$ and $n_r > 0$). % for $k \geq k_r$. % (in fact this isk^{p_a, p_0, \alpha}_{min, \BRAVO, n_r} = k^{*}_{n_r}$
% \textit{i.e.,}
% \begin{equation}\label{eq_incr}
% % \frac{1}{\alpha} \leq
%   \frac{P[K_r = {k_r} | H_a, n_r]}{P[\mathbf{k_r} | H_0, n_r]} \leq   \frac{P[\mathbf{k_r} + 1 | H_a, n_r]}{P[K_r = {k_r} + 1 | H_0, n_r]} \leq \ldots  \leq \frac{P[\mathbf{n_r} | H_a, n_r]}{P[\mathbf{n_r} | H_0, n_r]}. 
% \end{equation}

If \BRAVO test is satisfied for $k_r$ then:

% From the properties of binomial distribution we obtain:
% \[
%    \frac{P[\mathbf{k_r} | H_a, n_r]}{P[\mathbf{k_r} | H_0, n_r]} = 
%    \frac{P[\mathbf{k_{r-1}} | H_a, n_{r-1}]}{P[\mathbf{k_{r-1}} | H_0, n_{r-1}]}
%    \frac{P[\mathbf{k_r - k_{r-1}} | H_a, n_r - n_{r-1}]}{P[\mathbf{k_r - k_{r-1}} | H_0, n_r - n_{r-1}]} = 
% \]
\[
 \frac{1}{\alpha} \leq \sigma(k_{r}, p_a, p_0, n_{r}) =
\]
\[
 = \sigma(k_{r-1}, p_a, p_0, n_{r-1}) \cdot \sigma(k_r - k_{r-1}, p_a, p_0, n_r - n_{r-1})  \leq
\]
\[
 \leq^{(*)} \sigma(k_{r-1}, p_a, p_0, n_{r-1}) \cdot \tau_1(k_r - k_{r-1}, p_a, p_0, n_r - n_{r-1}) =
\]
\[
 = \omega_r(k_r, k_{r-1}, p_a, p_0, n_r, n_{r-1}).
\]

>>>>

To illustrate the practical implication of this property, we consider a toy example: an RLA of a two-candidate contest with margin $1\%$ and risk limit $10\%$. 
Suppose we wish to achieve a conditional stopping probability $0.9$ in each round of the audit. For \Providence, we can compute a new round size for each round based on the previous samples. For \Minerva, however, we would have a predetermined round schedule. We use the default \Minerva round schedule of audit software Arlo \cite{arlo} (used by most states performing an RLA), which is is $[x, 2.5x, 6.25x, ...]$; that is, the next marginal round size is $1.5$ times the current one. This multiplier of $1.5$ is known to give, over a wide range of margins, a probability of stopping roughly $0.9$ in the second round if the first round size has probability of stopping $0.9$.

Both the audits of our toy example therefore begin with a first round size of $17,272$ with a $0.9$ probability of stopping, and both will stop in the first round if the sample contains at least $8,725$ ballots for the winner. We now consider two cases for which the audit proceeds to a second round. 
\begin{description}
\item In one case there are $8,724$ votes for the winner in the sample, just one fewer than the minimum needed to meet the risk limit. In the \Minerva audit, we are already committed to a second round size of $43,180$ which, given the nearly-passing sample of the first round is higher than necessary, achieving a stopping probability in the second round of $.954$. The \Providence audit samples more than $9,000$ fewer ballots with a round size of $34,078$, achieving the desired $0.9$ probability of stopping.
\item In a less lucky sample, the winner recieves $8,637$ ballots, few more than the loser recieves. In the \Minerva audit, we again have to use a second round size of $43,180$, but now this round size only achieves a $0.727$ probability of stopping, significantly less than the desired $0.9$. Again, the \Providence audit can scale up the second round size according to the first sample and achieve the desired $0.9$ probability of stopping with $58,007$ ballots.
\end{description}

@@

\subsection{Efficiency}
\begin{lemma}
% For any round schedule %$[n_1, \ldots, n_r]$
For any risk-limit $\alpha \in (0, 1)$, for any margin
and for any round schedule $[n_1, \ldots, n_j]$, 
the \Providence RLA is more efficient than EoR \BRAVO.
\end{lemma}
\begin{proof}
 Let $[n_1, \ldots, n_j]$ be a round schedule, and lets assume that an EoR \BRAVO audit stopped exactly at
 round $j$, 
 after observing $k_1, \ldots, k_j$ ballots for the announced winner in each round respectively. This means that
 the following conditions hold:
 \begin{enumerate}
  \item for all $i = 1, \ldots, j-1$: 
%   $k_j < k^{p_a, p_0, \alpha}_{min, \BRAVO, n_j}$,
  $\frac{P[{K_i = k_{i}} | H_a, n_i]}{P[K_i = {k_{i}} | H_0, n_i]} < \frac{1}{\alpha}$,
  \item for $i = j$: % $k_r \geq k^{p_a, p_0, \alpha}_{min, \BRAVO, n_r}.$
  $\frac{P[{K_j = k_{j}} | H_a, n_j]}{P[K_j = {k_{j}} | H_0, n_j]} \geq \frac{1}{\alpha}$,
 \end{enumerate}

 From Lemma~\ref{lemma:tau1_increasing} 
 \[
 \tau_1(k, p_a, p_0, n_j) = 
  \frac{P[K_r \geq {k} | H_a, n_j]}{P[K_r \geq {k} | H_0, n_j]}
 \]
is a strictly increasing function of $k$ (for $0 < p_0, p_a < 1$ and $n_r > 0$). % for $k \geq k_r$. % (in fact this isk^{p_a, p_0, \alpha}_{min, \BRAVO, n_r} = k^{*}_{n_r}$
% \textit{i.e.,}
% \begin{equation}\label{eq_incr}
% % \frac{1}{\alpha} \leq
%   \frac{P[K_r = {k_r} | H_a, n_r]}{P[\mathbf{k_r} | H_0, n_r]} \leq   \frac{P[\mathbf{k_r} + 1 | H_a, n_r]}{P[K_r = {k_r} + 1 | H_0, n_r]} \leq \ldots  \leq \frac{P[\mathbf{n_r} | H_a, n_r]}{P[\mathbf{n_r} | H_0, n_r]}. 
% \end{equation}

If the \BRAVO test is satisfied for $k_j$ then:

% From the properties of binomial distribution we obtain:
% \[
%    \frac{P[\mathbf{k_r} | H_a, n_r]}{P[\mathbf{k_r} | H_0, n_r]} = 
%    \frac{P[\mathbf{k_{r-1}} | H_a, n_{r-1}]}{P[\mathbf{k_{r-1}} | H_0, n_{r-1}]}
%    \frac{P[\mathbf{k_r - k_{r-1}} | H_a, n_r - n_{r-1}]}{P[\mathbf{k_r - k_{r-1}} | H_0, n_r - n_{r-1}]} = 
% \]
\[
 \frac{1}{\alpha} \leq \sigma(k_{j}, p_a, p_0, n_{j}) =
\]
\[
 = \sigma(k_{j-1}, p_a, p_0, n_{j-1}) \cdot \sigma(k_j - k_{j-1}, p_a, p_0, n_j - n_{j-1})  
\]
\[
 \leq^{(*)} \sigma(k_{j-1}, p_a, p_0, n_{j-1}) \cdot \tau_1(k_j - k_{j-1}, p_a, p_0, n_j - n_{j-1}) =
\]
\[
 = \omega_r(k_j, k_{j-1}, p_a, p_0, n_j, n_{j-1}).
\]

### 2022-10-12 providence/sections/appendix.tex
toy example

\begin{lemma}
    \label{lemma:sigma_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\sigma(k,p_a,p_0,n)$ is strictly increasing as a function of $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
From $0<p_0< p_a< 1$, we get
$$
\frac{p_a}{p_0}>1
$$
and
$$
1-p_0>1-p_a\implies \frac{1-p_0}{1-p_a}>1,
$$
and thus
$$
\frac{p_a(1-p_0)}{p_0(1-p_a)}>1.
$$
Now simply observe that
$$
\frac{p_a(1-p_0)}{p_0(1-p_a)}\cdot\sigma(k, p_a, p_0, n)
=\frac{p_a(1-p_0)}{p_0(1-p_a)}\cdot\frac{p_a^{k} (1-p_a)^{n-k}}{p_a^{k} (1-p_a)^{n-k}}
$$$$
=\frac{p_a^{k+1} (1-p_a)^{n-(k+1)}}{p_a^{k+1} (1-p_a)^{n-(k+1)}}
=\sigma(k+1, p_a, p_0, n).
$$
\end{proof}

@@

\begin{proof}
    Note that $z_i$ is a weighted average of the values of $\frac{a_j}{b_j}$ for $j \geq i$:
    $$z_i = \sum_{j=i}^n y_j \frac{a_j}{b_j}$$  ~for~ $$y_j = \frac{b_j}{\sum _{j=i}^n b_j} > 0.$$
    Further, $\sum _{j=i}^n y_j = 1$ and hence  $y_j \leq 1$ ~and~ $y_j = 1 \Leftrightarrow i=j=n$.
    Observe that, because $\frac{a_i}{b_i}$ is monotone increasing,
    $z_i \geq \frac{a_i}{b_i}$
    with equality if and only if $i=n$.
    Suppose $i < n$. Then
    $$z_{i+1} \geq \frac{a_{i+1}}{b_{i+1}} > \frac{a_i}{b_i}, $$
    and
    $$z_i = y_i \frac{a_i}{b_i} + (1-y_i) z_{i+1} < z_{i+1}.$$
    Thus $z_i$ is also monotone increasing.
\end{proof}

\begin{lemma}
    \label{lemma:tau1_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\tau_1(k,p_a,p_0,n)$ is strictly increasing as a function $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
    Apply Lemmas \ref{lemma:sigma_increasing}-\ref{lemma:frac_sums_increasing}.
\end{proof}
\begin{lemma}
    \label{lemma:imin-exists}
    Given a strictly monotone increasing sequence: $x_1, x_2, \ldots x_n $ and some constant $A$,
    %there exists an $i_{min}$ such that
    $$A \le x_i \Leftrightarrow \exists i_{min} \le i ~\text{s.t.}~   x_{i_{min} -1} < A \le x_{i_{min}} \le x_{i},$$
    unless $A\le x_1$, in which case $i_{min} =1 $.
\end{lemma}
\begin{proof}
    Evident.
\end{proof}

\begin{lemma}
    \label{lemma:minerva2_kmin_exists}
    For $\mathcal{A}=(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence, there exists\\ a $k_{min,j}(\Providence, p_a, p_0, k_{j-1}, n_{j-1}, n_j)$ such that $$\mathcal{A}(X_j)=\text{Correct}\iff k_j\ge k_{min,j}(\Providence,  \bm{n_j}, p_a, p_0).$$
\end{lemma}
\begin{proof}
    From Definition~\ref{def:minervatwo}, $$\mathcal{A}(X_j)=\text{Correct}\iff \omega_j(k_{j}, k_{j-1}, p_a, p_0, n_j, n_{j-1}) \ge \frac{1}{\alpha}.$$
    Now to apply Lemma~\ref{lemma:imin-exists}, it suffices to show that
    $\omega_j$ is monotone increasing with respect to $k_j$.
    For $j=1$, we have $\omega_1=\tau_1$, so $\omega_1$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. For $j\ge 2$,
    $$\omega_j(k_j,k_{j-1},p_a,p_0,n_j,n_{j-1},\alpha)=$$$$\sigma(k_{j-1},p_a,p_0,n_{j-1})\cdot \tau_1(k_{j}-k_{j-1},p_a,p_0,n_j-n_{j-1}).$$
    As a function of $k_j$, $\sigma$ is constant, and thus $\omega$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. Therefore by Lemma \ref{lemma:imin-exists}, we have the desired property.
\end{proof}

@@

>>>>

\begin{lemma}
    \label{lemma:sigma_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\sigma(k,p_a,p_0,n)$ is strictly increasing as a function of $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
See \cite[Lemma 4]{usenix_minerva}. 
\end{proof}

@@

\begin{proof}
See \cite[Lemma 2]{usenix_minerva}. 
\end{proof}

\begin{lemma}
    \label{lemma:tau1_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\tau_1(k,p_a,p_0,n)$ is strictly increasing as a function of $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
    Apply Lemmas \ref{lemma:sigma_increasing}-\ref{lemma:frac_sums_increasing}.
\end{proof}
\begin{lemma}
    \label{lemma:imin-exists}
    Given a strictly monotone increasing sequence: $x_1, x_2, \ldots x_n $ and some constant $A$,
    %there exists an $i_{min}$ such that
    $$A \le x_i \Leftrightarrow \exists i_{min} \le i ~\text{s.t.}~   x_{i_{min} -1} < A \le x_{i_{min}} \le x_{i},$$
    unless $A\le x_1$, in which case $i_{min} =1 $.
\end{lemma}
\begin{proof}
    Evident.
\end{proof}

\begin{lemma}
    \label{lemma:minerva2_kmin_exists}
    For $\mathcal{A}=(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence, there exists\\ a 
    $k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  = 
    k_{min,j}(\Providence, p_a, p_0, k_{j-1}, n_{j-1}, n_j)$ such that $$\mathcal{A}(X_j)=\text{Correct}\iff k_j\ge k_{min,j}(\Providence,  \bm{n_j}, p_a, p_0).$$
\end{lemma}
\begin{proof}
    From Definition~\ref{def:minervatwo}, $$\mathcal{A}(X_j)=\text{Correct}\iff \omega_j(k_{j}, k_{j-1}, p_a, p_0, n_j, n_{j-1}) \ge \frac{1}{\alpha}.$$
    Now to apply Lemma~\ref{lemma:imin-exists}, it suffices to show that
    $\omega_j$ is monotone increasing with respect to $k_j$.
    For $j=1$, we have $\omega_1=\tau_1$, so $\omega_1$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. For $j\ge 2$,
    $$\omega_j(k_j,k_{j-1},p_a,p_0,n_j,n_{j-1},\alpha)=$$$$\sigma(k_{j-1},p_a,p_0,n_{j-1})\cdot \tau_1(k_{j}-k_{j-1},p_a,p_0,n_j-n_{j-1}).$$
    As a function of $k_j$, $\sigma$ is constant, and thus $\omega$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. Therefore by Lemma \ref{lemma:imin-exists}, we have the desired property.
\end{proof}

@@

\begin{definition} 
\label{def:kmin}
Let $[n_1, \ldots, n_j]$ be the round schedule of an audit that has not stopped by the round $j-1$. Let us define 
\begin{small}
\begin{equation}\label{eq:kMin}
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  =
  \min\left\{k : \omega_j(k, k_{j-1},p_a,p_0,n_j, n_{j-1}) \geq \frac{1}{\alpha}  \right\}.
%  \min\left\{k : \sigma(k_{r-1},p_a,p_0,n_{r-1})\cdot \tau_1(k-k_{r-1},p_a,p_0,n_r-n_{r-1}) \geq \frac{1}{\alpha}  \right\}$
\end{equation}
\end{small}
\end{definition}
As we have seen in Lemma \label{lemma:minerva2_kmin_exists}, such a value of $k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}$ exists and $k_j \geq k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $ if and only if the result of the audit is Correct, (\textit{i.e.,} the stopping condition in Definition~\ref{def:minervatwo} holds).

The following lemma shows a Markov-like property of \Providence audit (\textit{i.e.,}
for an audit that has not stopped in the first $j-1$ rounds, only cumulative results of the round $j-1$ matter: cumulative sample size $n_{j-1}$ and the number of ballots for the winner $k_{j-1}$).

\begin{lemma}\label{lemma:markov}
Let $[n_1, \ldots, n_{j-1}, n_j]$ be a round schedule for an execution of  \Providence audit that has not stopped
in any of its first $j-1$ rounds (\textit{i.e.,} for every $i = 1, \ldots, j-1$:
$k_i < k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $), then:

\[ 
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} = k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j} .
\]
\end{lemma}
% \fpo{this can be used to prove that \Providence is more efficient than \Minerva and \BRAVO}
\begin{proof}
Let $k_{j-1}$ denote the number of ballots drawn for the declared winner up to the round $j-1$ (out of $n_{j-1}$ sampled ballots). The stopping decision for the round $j$ is made as follows:

\[
 k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  = \min\left\{k : \omega_{j}(k, k_{r-1}, p_a, p_0, n_r, n_{r-1}) \geq \frac{1}{\alpha}  \right\} = 
\]
\[
  =  k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j}  
\]

\end{proof}

That is, the stopping condition is equivalent to that of a two round audit with the same cumulative votes for the winner and cumulative round sizes: the first round is of size $n_{j-1}$ and has $k_{j-1}$ votes for the winner, and the second (cumulative) round size is $n_j$ with $k_j$ (cumulative) votes for the winner. Compare this to the similar property for the $\Bravo$ stopping condition.
